Mental Rotation as Bayesian Quadrature

### 2013-11-27 man/nips-2013.tex
Uncomment \nipsfinalcopy

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
integration over a probability distribution, with the direction of
rotation becoming an optimal experiment design problem (or in machine
learning parlance, an active learning problem). We show that recent
work on methods for Bayesian quadrature
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm} provides a way to
solve this problem, outperforming several simpler heuristics for
determining both the direction and extent of rotation.

@@

For many years, this idea was contested, with some researchers arguing
the underlying cognitive processes were not visual in nature
(e.g., \cite{Pylyshyn1981}). In particular, Anderson
\cite{Anderson1978} proved that the mental imagery debate could not be
resolved on the basis of response time data alone. In recent years,
however, significant efforts have investigated the neural
underpinnings of the mental rotation phenomena, arguing that the
results of brain imaging studies support the existence of visual
imagery \cite{Kosslyn:2009tj,Kosslyn1988}.

We assume that mental imagery is indeed a visual process, and turn to
the question of how, and when, it is used. Previous work has examined
how people might use imagery \cite{Hegarty2004, Schwartz1999} to solve
reasoning problems, and when they might use imagery as opposed to a
symbolic rule \cite{Schwartz:1996uy}. Eye-tracking studies of people
performing mental rotations suggest strategies that people might be
using, such as aligning smaller features of the images
\cite{Just1976}. People seem to use imagery methodically, which raises
the question: what is the method?

@@

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method to place a prior over the log
likelihood\footnote{In practice, as in \cite{Osborne:2012tm}, we use
  the transform of $\log(S+1)$. Additionally, we follow
  \cite{Osborne:2012tm} and use a combination of MLII and
  marginalization to fit the $w$ kernel parameters. The output scales
  are fixed at $h_S=\sqrt{0.1}$ and $h_{\log S}=\log(h_S + 1)$.}, thus
ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have non-positive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points} $c$ as described in \cite{Osborne:2012tm}. The variance is
$\tilde{V}(Z\vert S, \log S, \Delta_c)$ as defined in Equation 12 of
\cite{Osborne:2012tm}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Top: each subplot shows the
    correspondence between the true angle of rotation ($R$) for
    ``same'' pairs and the amount of rotation performed by the
    model. Bottom: each subplot shows the models' mean rotations over
    stimuli pairs that were judged correctly. Black dots correspond to
    ``same'' pairs, and blue lines correspond to ``different'' pairs
    (for which the mean rotation is constant across true rotations, as
    the true rotation is undefined). Error bars/shaded regions
    indicate one standard deviation, and the dotted lines indicate the
    least-squares fit to the ``same'' pairs.}
  \label{fig:rotations}
\end{figure}

@@

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Response error rates}: How accurate was the model at
  choosing the correct hypothesis? This was defined as the mean error
  ($\ME{}$), or fraction of times the model chose incorrectly.
\item \textit{Rotations}: For those ``same'' pairs which the model
  judged correct, how correlated were the model's rotations with the
  true angles of rotation?  We quantified this using the Pearson's
  correlation coefficient $\rho$ for the true rotation ($R$) versus
  the extent of the model's rotations ($\vert \mathbf{R}\vert$). The
  top row in Figure \ref{fig:rotations} shows individual points
  corresponding to true rotations vs. rotations by the models for all
  stimuli, including those judged incorrectly. The bottom row shows
  mean model rotations, across only those stimuli which were judged
  correctly. We structured the analysis in this way to better compare
  to the results of \cite{Shepard1971}, which also excluded
  incorrectly judged stimuli.
\item \textit{Estimates of $Z$}: How accurate was the model's estimate
  of $Z$? We defined this quantity as the mean squared error
  ($\MSE{}$) between the model's estimate of $Z$ and the ``gold
  standard'' value, where the error has scale such that $\MSE{}=0$
  indicates no error and $\MSE{}=1$ indicates maximum error. Figure
  \ref{fig:accuracy} shows plots of the true (``gold standard'') value
  of $Z$ vs. the model's estimate, for each model.
\end{itemize}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. The \naive{}
    model (left panel) tends to overestimate. The parametric model
    (center panel) is very accurate for about half the ``same'' pairs,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true
    $Z$.}
  \label{fig:accuracy}
\end{figure}

@@

In this paper, we asked: how do people use mental simulation?  We
investigated the specific case of mental rotation, using rational
analysis to characterize optimal strategies for performing mental
rotation. We found that an adaptive, nonparametric model performs best
at the classic rotation task \cite{Shepard1971}, as it is able to
actively monitor the confidence of its estimate and intelligently
choose the direction of rotation. Two simpler models based on
heuristics performed much worse: they did not maintain the linear
relationship with the true angle of rotation due to lack of robustness
against local maxima or incorrect rotation direction. The
nonparametric model qualitatively approximates human behavioral
judgments, both in linearity \cite{Shepard1971} and variability
\cite{Just1976}. Future work will collect empirical data from
participants to perform a more thorough, quantitative analysis.  We
conclude that from this initial survey, models which take an
``active'', directed approach seem well-suited to explaining mental
simulation.

>>>>

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
integration over a probability distribution, with the direction of
rotation becoming an optimal experiment design problem (or in machine
learning parlance, an active learning problem). We show that recent
work on methods for Bayesian quadrature
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm} provides a way to
solve this problem, outperforming simpler heuristics for determining
the direction and extent of rotation.

@@

For many years, this idea was contested, with some researchers arguing
the underlying cognitive processes were not visual in nature (e.g.,
\cite{Pylyshyn1981}). In particular, Anderson \cite{Anderson1978}
proved that the mental imagery debate could not be resolved on the
basis of response time data alone. In recent years, however,
significant efforts have investigated the neural underpinnings of the
mental rotation phenomena, arguing that the results of brain imaging
studies support the existence of visual imagery
\cite{Kosslyn:2009tj,Kosslyn1988}.  We assume that mental imagery is
indeed a visual process, and turn to the question of how, and when, it
is used. Previous work has examined how people might use imagery
\cite{Hegarty2004, Schwartz1999} to solve reasoning problems, and when
they might use imagery as opposed to a symbolic rule
\cite{Schwartz:1996uy}. Eye-tracking studies of people performing
mental rotations suggest strategies that people might be using, such
as aligning smaller features of the images \cite{Just1976}. People
seem to use imagery methodically, which raises the question: what is
the method?

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.85\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Top: each subplot shows the
    correspondence between the true angle of rotation ($R$) for
    ``same'' pairs and the amount of rotation performed by the
    model. Bottom: each subplot shows the models' mean rotations over
    stimuli pairs that were judged correctly. Black dots correspond to
    ``same'' pairs, and blue lines correspond to ``different'' pairs
    (for which the mean rotation is constant across true rotations, as
    the true rotation is undefined). Error bars/shaded regions
    indicate one standard deviation, and the dotted lines indicate the
    least-squares fit to the ``same'' pairs.}
  \label{fig:rotations}
\end{figure}

Bayesian Quadrature has its difficulties, however. While in our case
$S$ is a non-negative likelihood function, GP regression enforces no
such constraint. In an effort to avoid this problem,
\cite{Osborne:2012tm} give a method to place a prior over the log
likelihood\footnote{In practice, as in \cite{Osborne:2012tm}, we use
  the transform of $\log(S+1)$. Additionally, we follow
  \cite{Osborne:2012tm} and use a combination of MLII and
  marginalization to fit the $w$ kernel parameters. The output scales
  are fixed at $h_S=\sqrt{0.1}$ and $h_{\log S}=\log(h_S + 1)$.}, thus
ensuring that $S=e^{\log S}$ will be positive\footnote{We are not
  guaranteed positivity, however, as the approximation to the integral
  over $\log S$ (Equation \ref{eq:bq-Z-mean}) requires computing
  $\mu_S$, which may have non-positive segments.}:
\begin{equation*}
  E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\ \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S}, \Sigma_{\log S}\right)\ \mathrm{d}\log S
\end{equation*}
where $\mu_{\log S}$ and $\Sigma_{\log S}$ are the mean and
covariance, respectively, of the GP regression over $\log S$ given
$\mathbf{R}$. We approximate this according to the method given in
\cite{Osborne:2012tm}:
\begin{equation}
  \mu_Z=E[Z\ \vert \ S, \log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\ \mathrm{d}R 
  \label{eq:bq-Z-mean}
\end{equation}
where $\mu_S$ is the mean of a GP regression over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ is a regression over
$\Delta_c=\mu_{\log S} - \log \mu_S$ given $\mathbf{R}_c$, which
consists of $\mathbf{R}$ and a set of intermediate \emph{candidate
  points} $c$ as described in \cite{Osborne:2012tm}. The variance is
$\tilde{V}(Z\vert S, \log S, \Delta_c)$ as defined in Equation 12 of
\cite{Osborne:2012tm}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.85\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. The \naive{}
    model (left panel) tends to overestimate. The parametric model
    (center panel) is very accurate for about half the ``same'' pairs,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true
    $Z$.}
  \label{fig:accuracy}
\end{figure}

@@

We considered three metrics of performance in particular:

\textit{Response error rates}: How accurate was the model at choosing
the correct hypothesis? This was defined as the mean error ($\ME{}$),
or fraction of times the model chose incorrectly.

\textit{Rotations}: For those ``same'' pairs which the model judged
correct, how correlated were the model's rotations with the true
angles of rotation?  We quantified this using the Pearson's
correlation coefficient $\rho$ for the true rotation ($R$) versus the
extent of the model's rotations ($\vert \mathbf{R}\vert$). The top row
in Figure \ref{fig:rotations} shows individual points corresponding to
true rotations vs. rotations by the models for all stimuli, including
those judged incorrectly. The bottom row shows mean model rotations,
across only those stimuli which were judged correctly. We structured
the analysis in this way to better compare to the results of
\cite{Shepard1971}, which also excluded incorrectly judged stimuli.

\textit{Estimates of $Z$}: How accurate was the model's estimate of
$Z$? We defined this quantity as the mean squared error ($\MSE{}$)
between the model's estimate of $Z$ and the ``gold standard'' value,
where the error has scale such that $\MSE{}=0$ indicates no error and
$\MSE{}=1$ indicates maximum error. Figure \ref{fig:accuracy} shows
plots of the true (``gold standard'') value of $Z$ vs. the model's
estimate, for each model.

@@

@@

In this paper, we asked: how do people use mental simulation?  We
investigated the specific case of mental rotation, using rational
analysis to characterize optimal strategies for performing mental
rotation. We found that an adaptive, nonparametric model performs best
at the classic rotation task \cite{Shepard1971}, as it is able to
actively monitor the confidence of its estimate and intelligently
choose the direction of rotation. Two simpler models based on
heuristics performed much worse: they did not maintain the linear
relationship with the true angle of rotation due to lack of robustness
against local maxima or incorrect rotation direction.  More broadly,
the nonparametric model provides answers to puzzling questions
surrounding the incremental nature of mental rotation: which way
should the object be rotated, and for how long? This model formalizes
these answers in a way that is qualitatively consistent with human
behavior, both in response time linearity \cite{Shepard1971} and
variability \cite{Just1976}. These results further indicate that this
may be another instance in which people do appropriately use available
computational resources to solve the task at hand.  Rational, adaptive
approaches such as this may therefore be successful when applied to
more complex cognitive processes above and beyond mental rotation.
